---
title: "A3"
author: "Tian Chen"
date: "`r Sys.Date()"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Set up environment and import data

```{r}
#clear environment
rm(list = ls())

#import package
require(tidyverse)
#import data
datsss <- read.csv("~/Desktop/Duke study/Econ613/A3/datsss.csv")
datjss <- read.csv("~/Desktop/Duke study/Econ613/A3/datjss.csv")
datstu_v2 <- read.csv("~/Desktop/Duke study/Econ613/A3/datstu_v2.csv")
datsss_tbl <- datsss %>% as_tibble() #school
datjss_tbl <- datjss %>% as_tibble() #geolocation of school
datstu_v2_tbl <- datstu_v2 %>% as_tibble() #students

```

## Exercise 1
###1. Number of Students, schools, programs

```{r}
#Number of students
length(unique(datstu_v2$V1))

#Number of Schools
length(unique(datsss$schoolname))

#Number of programs
program <- select(datstu_v2_tbl, choicepgm1:choicepgm6)
program_name <- names(program)
unique_name <- vector()

#Find unique program name
for (i in c(1:length(program_name))) {
  unique_col <- unique(select(program, program_name[i]))
  unique_col <- rename(unique_col, program = program_name[i])
  unique_name <- unique(rbind(unique_name, unique_col))
}
length(unique_name$program)
```

###2. Number of Choices
```{r}
school_program <- datstu_v2_tbl %>% select(V1, schoolcode1:choicepgm6)
#transfer from wide to long
program<- school_program %>% select(V1, choicepgm1:choicepgm6) %>% pivot_longer(cols = starts_with("choice"),
                                                  names_to = "choice_progran",
                                                  values_to = "program_name") 
school <- school_program %>% select(V1, schoolcode1:schoolcode6) %>% pivot_longer(cols = starts_with("schoolcode"),
                                                  names_to = "choice_school",
                                                  values_to = "schoolcode")

school_program <- cbind(school, program) %>% select(program_name, schoolcode) %>% unique()
length(school_program$schoolcode)
```

###3. Number of students applying to at least one senior high school in the same district to home
```{r}
students_school <- datstu_v2_tbl %>% select(V1, schoolcode1:schoolcode6, jssdistrict) %>% pivot_longer(cols = starts_with("schoolcode"),
                                                                                                       names_to = "choice_school",
                                                                                                       values_to = "schoolcode")
school_location <- datsss_tbl %>% select(schoolcode, sssdistrict)
students_school <- left_join(students_school, school_location, by = "schoolcode") 
students_school <- students_school %>% mutate(same_district = ifelse(students_school$jssdistrict == students_school$sssdistrict, 1, 0))
#count the at least one same location
students_school <- students_school %>% group_by(V1) %>% summarise(count = sum(same_district)) %>% filter(count>=1)
length(students_school$V1)
```

###4. Number of students each senior high school admitted
```{r}
school_number <- datstu_v2_tbl %>% select(V1, schoolcode1:schoolcode6, rankplace) %>% drop_na() %>% pivot_longer(cols = starts_with("schoolcode"),
                                                                                                       names_to = "choice_school",
                                                                                                       values_to = "schoolcode")

school_number <- school_number %>% mutate(choice_school_number = 0, 
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode1", 1),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode2", 2),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode3", 3),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode4", 4),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode5", 5),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode6", 6),
                                              )
school_number <- school_number %>% mutate(admitted = ifelse(school_number$choice_school_number == school_number$rankplace, 1, 0))
school_number <- school_number %>% select(schoolcode, admitted) %>% group_by(schoolcode) %>% summarise(count = sum(admitted))

```

###5. The cutoff of senior high schools 
```{r}
school_score <- datstu_v2_tbl %>% select(V1, score, schoolcode1:schoolcode6, rankplace) %>% drop_na() %>% pivot_longer(cols = starts_with("schoolcode"),
                                                                                                       names_to = "choice_school",
                                                                                                       values_to = "schoolcode")

school_score <- school_score %>% mutate(choice_school_number = 0, 
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode1", 1),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode2", 2),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode3", 3),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode4", 4),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode5", 5),
                                              choice_school_number = replace(choice_school_number, choice_school == "schoolcode6", 6),
                                              )
school_score <- school_score %>% mutate(admitted = ifelse(school_score$choice_school_number == school_score$rankplace, 1, 0)) %>% filter(admitted == 1)

school_cutoff <- school_score %>% select(score, schoolcode) %>% group_by(schoolcode) %>% summarise(cutoff = min(score))
head(school_cutoff)
```

###6. The quality of senior high schools 
```{r}
school_quality <- school_score %>% select(score, schoolcode) %>% group_by(schoolcode) %>% summarise(quality = mean(score))
head(school_quality)
```

## Exercise2 Data
```{r}
#(school, program)
#transfer from wide to long
program<- datstu_v2_tbl %>% select(V1, choicepgm1:choicepgm6) %>% pivot_longer(cols = starts_with("choice"),
                                                  names_to = "choice_progran",
                                                  values_to = "program_name") %>% select(!V1)
school <- datstu_v2_tbl %>% select(!(choicepgm1:choicepgm6)) %>% pivot_longer(cols = starts_with("schoolcode"),
                                                  names_to = "choice_school",
                                                  values_to = "schoolcode")
#find cutoff, quality, and size
school_program <- cbind(program, school) %>% mutate(choice_school_number = 0, 
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode1", 1),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode2", 2),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode3", 3),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode4", 4),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode5", 5),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode6", 6),
                                              )
school_program <- school_program %>% mutate(admitted = ifelse(school_program$choice_school_number == school_program$rankplace, 1, 0))%>% filter(admitted == 1)
school_program <- school_program %>% group_by(schoolcode, program_name) %>% summarize(cutoff = min(score, na.rm = TRUE),
                                                                        quality = mean(score, na.rm = TRUE),
                                                                        size = n())

#merge (school, program) with datsss
school_info <- datsss_tbl %>% select(-V1) %>% drop_na() %>% unique() %>% mutate(length = str_count(schoolname))
#because of the data duplication, we use the longest name in each schoolcode as the school name
school_info_unique <- school_info %>% group_by(schoolcode) %>% slice_max(length, n = 1) %>% select(-length)
school_level_data <- left_join(school_program, school_info_unique, by = "schoolcode")
head(school_level_data)

```

## Exercise3 Distance

```{r}
#individual school_program data
ind_data <- cbind(school, program)
#merge with school-level data
ind_data <- left_join(ind_data, school_level_data, by = c("program_name", "schoolcode"))
#merge with datjss (home location)
location <- datjss_tbl %>% drop_na() %>% filter(jssdistrict != "")
ind_withdistance <- left_join(ind_data,location, by = "jssdistrict") %>% rename(jsslong = point_x, jsslat = point_y)
#calculate the distance 
ind_data_withdistance <- ind_withdistance %>% mutate(distance = sqrt((69.172*(ssslong-jsslong)*cos(jsslat/53))^2 + (69.172*(ssslat-jsslat))^2))
head(ind_data_withdistance)
```

##Exercise4 Dimensionality Reduction

###Recode the schoolcode into its three digits(substr). Call this new variable scode_rev
```{r}

data <- ind_data_withdistance
data <- data %>% mutate(scode_rev = substr(schoolcode, 1, 3))
```

###Recode the program variable into 4 categories. Call this new variable pgm_rev
```{r}
#Recode the program variable into 4 categories 
data <- data %>% mutate(pgm_rev = ifelse(data$program_name == "General Arts"|data$program_name == "Visual Arts", "arts",
                                         ifelse(data$program_name == "Business"|data$program_name == "Home Economics", "economics",
                                                ifelse(data$program_name == "General Science", "science", "others"))))
```

###Create a new choice variable choice_rev
```{r}
data <- data %>% mutate(choice_rev = paste(data$scode_rev, data$pgm_rev))
```

###Recalculate the cutoff and the quality for each recoded choice
```{r}
cutoff_quality<- data %>% mutate(choice_school_number = 0, 
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode1", 1),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode2", 2),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode3", 3),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode4", 4),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode5", 5),
                                            choice_school_number = replace(choice_school_number, choice_school == "schoolcode6", 6),
                                              )
cutoff_quality <- cutoff_quality %>% mutate(admitted = ifelse(cutoff_quality $choice_school_number == cutoff_quality $rankplace, 1, 0))%>% filter(admitted == 1)
cutoff_quality <- cutoff_quality %>% group_by(choice_rev) %>% summarize(cutoff = min(score,na.rm = TRUE),
                                                                        quality = min(score, na.rm = TRUE))
data <- data %>% select(-cutoff, -quality) %>% left_join(cutoff_quality, by = "choice_rev")
```

###Consider the 20,000 highest score students
```{r}
top_20000_students <- data %>% select(V1, score) %>% unique() %>% arrange(desc(score)) 
data_top20000 <- data %>% filter(V1 %in% top_20000_students$V1[1:20000])
head(data_top20000)
```

##Exercise 5 First Model

###Using the new data with recoded choices, we want to understand the effect of the student test score on his first choice.
###Propose a model specification. Write the likelihood function.

###likelihood function

```{r}
#I decided to use the conditional logit model to estimate the impacts of test scores on the probability of first choice
## create a data set
test_data <- data_top20000 %>% select(score, choice_rev, choice_school)
test_data <- test_data %>% filter(choice_school == "schoolcode1") %>%mutate(choice=as.numeric(factor(choice_rev,ordered=TRUE))) %>% drop_na()
data <- test_data
##likelihood function
like_fun = function(guess)
{
  score =  data$score
  choice = data$choice
  
  row = nrow(data)
  column = length(unique(choice))
  ut = mat.or.vec(row,column)
  
  for (j in 1:column)
  {
    ut[,j] = guess[1] + guess[2]*score[j]
  }
  prob   = exp(ut)
  prob   = sweep(prob,MARGIN=1,FUN="/",STATS=rowSums(prob))
  prob2 = prob[,1]
  prob2[prob2>0.999999] = 0.999999
  prob2[prob2<0.000001] = 0.000001
  like = sum(log(prob2))
  return(-like)
}


#we guess the parameter, set the second one as negative
guess <- runif(2) 
guess[2] <- -guess[2]
like_fun(guess = guess)


#simulation
optim(par = guess,fn=like_fun)

#simulation
#optim(par = guess,fn=like_fun)
```


###Question 6

###Using the new data with recoded choices, we want to understand the effect of the school_quality on his first choice.
###Propose a model specification. Write the likelihood function. 


```{r}
#I decided to use the conditional logit model to estimate the impacts of school quality on the probability of first choice
## create a data set
test_data <- data_top20000 %>% select(quality, choice_rev, choice_school)
test_data <- test_data %>% filter(choice_school == "schoolcode1") %>%mutate(choice=as.numeric(factor(choice_rev,ordered=TRUE))) %>% drop_na()
data <- test_data
##likelihood function
like_fun = function(guess)
{
  quality =  data$quality
  choice = data$choice
  
  row = nrow(data)
  column = length(unique(choice))
  ut = mat.or.vec(row,column)
  
  for (j in 1:column)
  {
    ut[,j] = guess[1] + guess[2]*quality[j]
  }
  prob   = exp(ut)
  prob   = sweep(prob,MARGIN=1,FUN="/",STATS=rowSums(prob))
  prob2 = prob[,1]
  prob2[prob2>0.999999] = 0.999999
  prob2[prob2<0.000001] = 0.000001
  like = sum(log(prob2))
  return(-like)
}


#we guess the parameter, set the second one as negative
guess <- -runif(2) 
like_fun(guess = guess)

#simulation
optim(par = guess,fn=like_fun)
```

###Question 7 Counterfactual simulations

##construct the data
```{r}
test_data <- data_top20000 %>% select(quality, choice_rev, choice_school)
test_data <- test_data %>% filter(choice_school == "schoolcode1") %>%mutate(choice=as.numeric(factor(choice_rev,ordered=TRUE))) %>% drop_na()
data <- test_data
```
##Which model is proper?
```{r}
#Schools' quality may affects the types of program they can provide to students. When excluding the other programs, school's quality should be less correlative to the choice outcome. Thus, we should use the first model
```

##Simulate the model
```{r}

like_fun = function(guess)
{
  score =  data$score
  choice = data$choice
  
  row = nrow(data)
  column = length(unique(choice))
  ut = mat.or.vec(row,column)
  
  for (j in 1:column)
  {
    ut[,j] = guess[1] + guess[2]*score[j]
  }
  prob   = exp(ut)
  prob   = sweep(prob,MARGIN=1,FUN="/",STATS=rowSums(prob))
  prob2 = prob[,1]
  prob2[prob2>0.999999] = 0.999999
  prob2[prob2<0.000001] = 0.000001
  like = sum(log(prob2))
  return(-like)
}


#we guess the parameter, set the second one as negative
guess <- runif(2) 
guess[2] <- -guess[2]
like_fun(guess = guess)


#simulation
optim(par = guess,fn=like_fun)
```

